Consistency and Fault Tolerance Considerations for the Next Iteration of the DOE Fast Forward Storage and IO Project

### 2014-03-15 paper.tex
revised according to conversation with Eric Barton

Current production HPC IO stack design is unlikely to offer sufficient features
and performance to adequately serve the needs of an extreme scale platform. To
address these limitations, the US Department of Energy commissioned an effort
to develop a design and prototype for an IO stack suitable for the extreme
scale environment. This is a joint effort led by the Intel Lustre team, Los
Alamos National Laboratory, EMC, and the HDF Group. This team has developed a
specification set~\cite{fastforward:2014:docs} for a future IO stack to address
the identified challenges. The first phase recently completed with a second
phase getting underway. The core focus of the first phase was basic
functionality and design. The second phase will refine this design
incorporating fault recovery and other features missing from the first phase.

@@

Offloading the collective two-phase data sieving from the compute nodes to
reorganize data has proven effective at reducing the total time for writing
data due to fewer participants involved in the communication
patterns~\cite{lofstead:2011:nessie-staging}.  Beyond these broad items, there
are many important details some of which are examined in more detail below.

@@

%Incorporated into the design of these layers and the function of some of these
%layers themselves, such as the IOD layer, are features to manage both burst IO
%performance and faults. In particular, the IOD layer's use of burst buffers
%attempts to absorb a massive data dump from the compute area that is then
%trickled to the persistent storage area only if desired. Otherwise, it can be
%stored locally on optional SSDs for use by analysis routines eliminating the
%need to use slower rotational media as the intermediary between a simulation
%and the analysis code.

%A second feature is the incorporation of a transaction like mechanism into the
%IOD layer and a related epochs feature into the DAOS layer. This feature offers
%a mechanism to avoid making incomplete data visible to other users and offer a
%level of protection if the data can be stored on the optional SSDs. The
%implications of how these concepts are designed are important to consider for
%the final design. Because they control data visibility and manage faults for
%output operations, they are key features.

@@

The idea of burst buffers were initially explored in the context of data
staging~\cite{abbasi:2007:datatap,Abbasi:2009:datatap,nisar:2008:staging,zheng:2010:predata}.
These initial designs all use extra compute nodes to represent the data storage
buffer given the lack of any hardware support for this functionality. The
desired outcome of these initial studies is to motivate how such functionality
might be incorporated and the potential benefits.  Later, these concepts were
proposed to be incorporated as part of the IO
stack~\cite{bent:2012:challenges,bent:2012:burst-buffer}.  The current
FastForward IOD design recommends incorporating SSDs, but specifically lists
these devices as optional. Unfortunately, both incorporating burst buffers and
the use of SSDs in the IOD layer may be problematic.  First, with these burst
buffers being optional, the semantics of how IOD works must change to use DAOS
to store the data directly rather than storing them locally until explicitly
persisted by the user. The design of DAOS does not incorporate a high
transaction rate mechanism. Conversely, it assumes that it will only be
involved when persisting a completed transaction and only for a fraction of the
total transactions created. Further, with function shipping offering the
ability to change how the data is stored and arranged prior to it being written
to DAOS, this functionality may also be optional depending on the existence of
the burst buffer. Consider the important functionality of data rearrangement
and doing things like changing the fast array dimension.

@@

In the case of the IOD design, it forces a staging area deployment and then
shares the staging area across all users simultaneously. This is unlikely to be
useful because of the limited compute and communication capacity to spare to
perform these operations at a bottleneck in the IO path. The use of a separate
data staging area intentionally separate from the IO path allows using
operators on limited resources leaving the IO path clear for strictly data
movement.

@@

@@

@@

\subsection{Metadata Management}
While a central design goal of the Fast Forward project to eliminate the
requirement for a metadata server as a core component, it was not fully
accomplished.  This idea is not new and has been explored as part of the Light
Weight File Systems project~\cite{oldfield:lwfs}. In LWFS, the metadata service
is explicitly limited to a user task with the storage layer limited to data
storage/retrieval, authorization, and authentication. This approach proved
workable. For the Fast Forward project, while a similar goal is desired, it is
not completely accomplished.

The purpose behind the goal is to eliminate the serialized bottleneck caused by
having a centralized metadata service. Much of the design of IOD with objects
and containers does work to eliminate these bottlenecks.  However, IOD does not
go fully to a no metadata service model. Instead, the IOD layer describes a
key-value metadata service it provides. This service maintains four different
items:

\begin{itemize}
\item The list of IOD objects within the container
\item The mapping from each IOD object to the corresponding DAOS object.
\item The sharding and striping of the IOD object
\item The maximum valid offset of the object
\end{itemize}

While a pure no metadata model would be ideal for performance, this current
proposal has a few challenges. By providing this basic functionality, metadata
management is split between the user layer and the IOD layer. Consider the
following items. First, there is no list of containers. This is closer to the
no metadata ideal, but given the rest of the functionality, it exports to the
user level a requirement to track what containers are in the file system.
Second, given the flattening that occurs between IOD and DAOS, the mapping from
IOD objects to DAOS objects is unlikely to be 1-to-1. This is likely a list
instead. Further, given the ability to re-stripe in the IOD layer, the list is
not a list of objects completely included in the IOD object, but instead
objects from which part of the data came from some listed DAOS object.  Third,
sharding IOD objects makes sense based on the compute process connection a
shard has with the complete object.  Striping of an IOD object makes sense in
connection with the DAOS object(s) underneath, but striping within the IOD
layer is not discussed. Further, using striping from clients into the IOD layer
introduces a level of coordination that IOD seeks to avoid.  Fourth, given the
previous items, there is ambiguity with the maximum valid offset of the object.
This could be for the shard or for the collection of shard objects that
represent a distributed variable. Ideally, either a robust metadata management
system should be incorporated into the system, preferably separate from the IOD
layer or all of this functionality should be left to the user layer. The only
potential challenge is determining the container for a file open command and
what the transaction ID should be. This limited operation should be a
straightforward search using a hash or similar fast mechanism for examining a
list of container names and/or object IDs.

@@

For the IOD proposal, the service needs some adjustments to be generally
useful.  First, there needs to be a way to query the list of available
containers. This will support an analysis application using data generated by
the simulation code. This use case is a stated motivating example for the Fast
Forward project. Second, the maximum transaction ID for a container is
mentioned as being maintained by the IOD metadata service, but when and how
this is done is never mentioned.  Third, the definition of an object needs to
be made more clear. In some cases, it seems that an object is a globally
distributed object is represented by a collection of shards. In other cases, it
is a single shard.

@@

First, \DDT has a timeout mechanism to detect failures and offer an ability to
recover and clean up the incomplete transactions. IOD's transactions are
asynchronous precluding a relatively simple, short timeout value to detect
failures. Unfortunately, at this point, there is no mechanism in IOD's
transaction handling to detect a fault and clean up any incomplete operations.
While this is intended for phase 2, some initial suggestion on how this might
be handled would give some confidence that the asynchronous approach is viable.

Second, should a transaction on a container not complete, all subsequent
transactions on that container, even if they are complete and correct, will not
be accessible.

Third, in the single leader model, if the process that manages the transaction
were to fail, there is no ability to abort that transaction. The data will not
be made available because the transaction is still listed as in progress. The
problem is that without a way to clean up this failed transaction, no newer
versions of the same container can be made readable. In reality, the single
leader model would ideally outsource the transaction processing to something
like \DDT on the client side. This does not address the failure case though.

Fourth, in the multi-leader model, using a count of client connections to
determine if transaction is complete is problematic.  First, should this be in
an environment where lost or repeated network messages are possible, then the
count could be wrong inappropriately. This is a general problem and unlikely
in the current environment, but is a concern on less integrated platforms.
The idea of a count precludes any tracking of expected messages from sources
introducing a potential hole. Second, the aggregation of completion messages
are based on the local aggregation point reporting to the transaction
leader the state.  There is no ability to detect or recover from a failure of
this IOD process that would report to the transaction leader. Third, because a
simple count of participating processes is sent with the beginning of the
transaction, each process is strictly limited to a single output operation.
This precludes a process writing multiple array shards, for example. To put
this in concrete terms, it means that a file could only contain a single,
globally distributed variable (at the root level of HDF5) OR an attribute per
process. If this count actually represents a begin/end transaction pair
instead, it suffers from a lack of accountability for missing object (shard)
writes. All it would indicate is that a process started and ended a transaction
connection. There is no validation that a process has done everything required
of it.

@@

Consider a shared file system across an HPC data center. The current design
maintains the metadata in the IOD layer localized to a single machine
effectively making the data inaccessible from another platform. This requires
either the IOD layer be shared across machines, something clearly beyond the
design scope, or that an additional metadata service that extends the minimal
metadata service provided in the IOD layer. This higher level metadata service
would be required to be reachable from any machine that shares the file system.
As explained above, having this client-side only metadata service introduces
additional functionality requirements.

>>>>

Current production HPC IO stack design is unlikely to offer sufficient features
and performance to adequately serve the needs of an extreme scale platform. To
address these limitations, the US Department of Energy commissioned an effort
to develop a design and prototype for an IO stack suitable for the extreme
scale environment. This is a joint effort led by the Intel Lustre team, Los
Alamos National Laboratory, EMC, DDN, and the HDF Group. This team has
developed a specification set~\cite{fastforward:2014:docs} for a future IO
stack to address the identified challenges. The first phase recently completed
with a second phase getting underway. The core focus of the first phase was
basic functionality and design. The second phase will refine this design
incorporating fault recovery and other features missing from the first phase.

The overall design seeks to offer a byte-granular, multi-version concurrency
control. Through the use of a copy-on-write style mechanism, multiple versions
of a file can be stored in potentially greatly reduced space. By assuming the
client interface will be through an IO library affords a more complicated
interface that offers richer functionality without requiring user code changes.
By managing data in the IOD layer rather than requring writing to centralized
storage will better support the performnce requirements of integrated
application workflows. At a more detailed view, the various layers of the IO
stack each contribute different functionality and performance implications.

@@

While these ideas are not necessarily new, they are new twists on best of class
efforts for these techologies. For example, offloading the collective two-phase
data sieving from the compute nodes to reorganize data has proven effective at
reducing the total time for writing data due to fewer participants involved in
the communication patterns~\cite{lofstead:2011:nessie-staging}.  Beyond these
broad items, there are many important details some of which are examined in
more detail below.

@@

Along with the analysis of the published design documents, a discussion of the
design philosophy representing the overall intent is presented. This informtion
reprents information that may or moy not have been written down, but is the 
intent of ultimate product. These insights were gained based on personal
converations with the team members discussing some of the potential challenges
with the design as written. These ideas are presented to give a fuller picture
of where the project is going rather than dwelling on the limitations of the
published documents.

@@

The idea of burst buffers were initially explored in the context of data
staging~\cite{abbasi:2007:datatap,Abbasi:2009:datatap,nisar:2008:staging,zheng:2010:predata}.
These initial designs all use extra compute nodes to represent the data storage
buffer given the lack of any dedicated hardware support for this functionality.
The desired outcome of these initial studies is to motivate how such
functionality might be incorporated and the potential benefits.  Later, these
concepts were proposed to be incorporated as part of the IO
stack~\cite{bent:2012:challenges,bent:2012:burst-buffer}.  The current
FastForward IOD design recommends incorporating SSDs, but specifically lists
these devices as optional. Unfortunately, both incorporating burst buffers and
the use of SSDs in the IOD layer may be problematic.  First, with these burst
buffers being optional, the semantics must be clarified of how IOD works must
change to use DAOS to store the data directly rather than storing them locally
until explicitly persisted by the user. The design of DAOS does not incorporate
a high transaction rate mechanism. Conversely, it assumes that it will only be
involved when persisting a completed transaction and only for a fraction of the
total transactions created. Further, with function shipping offering the
ability to change how the data is stored and arranged prior to it being written
to DAOS, this functionality may also be optional depending on the existence of
the burst buffer or slow the IO path to accommodate the data processing.
Consider the important functionality of data rearrangement and doing things
like changing the fast array dimension.

@@

In the case of the written IOD design, it forces a staging area deployment and
then shares the staging area across all users simultaneously. This is unlikely
to be useful because of the limited compute and communication capacity to spare
to perform these operations at a bottleneck in the IO path. The use of a
separate data staging area intentionally separate from the IO path allows using
operators on limited resources leaving the IO path clear for strictly data
movement.

@@

\subsection{Design Philosophy}
The burst buffers design, as presented in the IOD documents, limits the
placement of the function operators and SSDs for buffers to the IO nodes. The
team does acknowledge the limitations of this design and intend to ultimately
focus on spreading the IOD from the IO nodes into the compute area as well.
This is intended both to help address the limitations of the IO bandwidth
and compute capaibility of these few nodes for data processing, but also to
take advantage of new layers in the storage hierarchy. By incorporating NVRAM
into compute nodes, new options for buffering data prior to being moved to
centralized storage become available. This lessens the impact of some operators
while offering additional options for places to store data.

Burst buffers being optional is a high level goal, but not one considered at a
detailed level with the design. While this forces some serious thought on how
to adapt the functionality proposed for IOD to work directly on DAOS will have
to be considered. With the additional desire to support using compute node
resources for these operations, serious work will be required to make a fully
functional end-to-end IOD layer implementation for a production system.

Another concern that is acknowledged, but no thought as been applied to is the
requirement that a single IOD process be the master for any operation. Should
the number of concurrent applications exceed the available nodes, sharing an
IOD process will be required. Since partitioning of the IOD processes for
exclusive use by particular applications is the proposed operating mode, should
insufficient IOD resources be available, either a job could be delayed or IOD
resources could be reallocated from a different process could be redeployed for
use by the new job. These sorts of consideraitons still need to be made for a
full production system.

@@

\subsubsection{Design Philosophy}
Undocumented, but inherent in the design of these transactions is how faults
are detected. The initial design assumes that the current Lustre fault
detection mechanism that can determine if a process or node is no longer
reachable is used. This defines how a fault will be detected and what will trigger a passive fault recovery (i.e., transaction abort).

A new term, ``Primary'', is used to define the process(es) that contact the IOD
layer directly to open a container. Unfortunately, this term is never used nor
defined within any of the published documents. There are some very positive and
negative performance implications of this approach. On the positive side, like
PVFS, it is possible for a single process to open a container and share that
handle with other processes reducing the load on the IOD layer. The key idea is
that the single leader vs. multi-leader transaction mode is automatically defined based on the number of Primary processes that open a container. This greatly
simplifies any additional client-side load in using the different styles of transactions. Also on the positive side, it offers the ability to have a subset of
processes that are considered transaction participants affording reduced load
on the IOD processes and more localized transaction management on the client
side. On the negative side, for a fully IOD managed transaction, ALL processes
must open the container forcing a potentially immense number of simultaneous
calls to the IOD layer to open a container. This sort of onslaught is what is
dealt with on metadata servers today with mass file opens. The reality is that
this ends up having the same sort of performance implications as distributed
denial of service attacks on the IOD processes.

Ultimately, with the passive detection of faults for Primaries, the transaction
mechanism can work very well. A mostly unstated restriction that will be
relaxed in the next phase is that every sequential transaction on a container
is considered dependent on the earlier transaction. Should one output be
delayed and the subsequent five succeed, when the delayed process finally
fails, all six transactions are rolled back. The thought of using this
mechanism to store subsequent checkpoint outputs in the same container to both
save space, but not care if one fails, cannot work in the current form. This
has been acknowledged and is planned on being relaxed in phase two.

@@

\subsection{Metadata Management}
Eliminating metaata management as a special case and instead treating it just
as data is a central design goal of the Fast Forward project. Unfortunately,
it is not completely achieved.

Eliminating metadata as a core component of a file system is not new. It has
been explored as part of the Light Weight File Systems
project~\cite{oldfield:lwfs}. In LWFS, the metadata service is explicitly
limited to a user task with the storage layer limited to data
storage/retrieval, authorization, and authentication. This approach proved
workable. Using this hybrid approach is less common and introduces other
issues.

IOD and DAOS both share a philosophy that they will have to maintain the
metadata about how objects are striped and where they are placed. It was
decided that the additional data to track the namespace as well was a small
enough additional increment that it was worth maintaining within the system.

\subsubsection{Design Philosophy}
While the metadata design is not fully defined, there are a few things that
are intended as part of this. For example, there is a standard, well known
container that is the system metadata. This includes the list of all other
containers. This container is treated like any other data in the system and
striped as appropriate at both the IOD and the DAOS layers. Unfortunately, this
still couples the metadata to a single object that must serialize access. If
the metadata, including information about striping and other data layout
operations were separated completely from the data path, more scalable
throughput could be achieved. The real challenge of this is introduced by the
the IOD, DAOS, and VOSD layers. Each of these requires some different metadata
storage and the migration is transparent to the user. Supporting fully
independent metadata with this model is difficult, if not impossible. Serious
thought on how to do this outside the data path effectively should be
considered for phase two.

@@

@@

First, should a transaction on a container not complete, all subsequent
transactions on that container, even if they are complete and correct, will not
be accessible. Once the transaction fails, all of these subsequent transactions
will be aborted. While this is being reconsidered for phase two, it does
complicate the interface to offer the additional transaction mode.

Second, in the multi-leader model, using a count of client connections to
determine if transaction is complete is problematic.  First, should this be in
an environment where lost or repeated network messages are possible, then the
count could be wrong inappropriately. This is a general problem and unlikely
in the current environment, but is a concern on less integrated platforms.
The idea of a count precludes any tracking of expected messages from sources
introducing a potential hole. Second, the aggregation of completion messages
are based on the local aggregation point reporting to the transaction
leader the state.  There is no ability to detect or recover from a failure of
this IOD process that would report to the transaction leader. Third, because a
simple count of participating processes is sent with the beginning of the
transaction, each process is strictly limited to a single output operation.
This precludes a process writing multiple array shards, for example. To put
this in concrete terms, it means that a file could only contain a single,
globally distributed variable (at the root level of HDF5) OR an attribute per
process. If this count actually represents a begin/end transaction pair
instead, it suffers from a lack of accountability for missing object (shard)
writes. All it would indicate is that a process started and ended a transaction
connection. There is no validation that a process has done everything required
of it.

@@

At a broader level, there are some concerns that were partially clarified
through conversations with the team.  Consider a shared file system across an
HPC data center. The current design maintains the metadata in its own
container. Since copying data from the IOD layer to the DAOS layer requires an
explicit persist call, how and when synchronizing the metadata across the 
layers and potentially across machines occurs is undefined. By delaying the
synchronization until an explicit persist is called will reduce the update
frequency, it does delay the visibility of data. Ideally, the metadata object
would need to be automatically persisted every time a container transaction is
persisted to the DAOS layer.

The implication of this is that every transaction persist is double operation
to account for the metadata persist. More importantly, the IOD-layer version of
the metadata container may contain readable transactions that have not been
persisted to the DAOS layer. How to handle this inconsistency between the two
layers still needs to be explored.
